\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2017-05-06 manuscript.tex

Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure. There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and other Type 2. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates. Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.  Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable. Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.

>>>>

Despite the power of the LSM in representing network structure that is not modeled by the observed covariates, and its performance in predictive ties predicting ties, we have reason to doubt the LSM's capacity to adjust for confounding network structure. We do not claim that the methodological literature in which the LSM was introduced or extended has presented the LSM as capable of adjusting for unmeasured confounding structure. However, as we document below, many researchers who have applied the LSM claim that it has the capacity to adjust for confounders, and use it for this purpose. There are two reasons that using the LSM may lead to increased error---one may result in Type 1 inferential error, and the other in Type 2 error. First, the latent configurations inferred may result in a representation of the network wherein a node's position in the latent space is spuriously correlated with the observed covariates. Inferring latent variables that are correlated with observed covariates would lead to reduced efficiency in estimating observed covariate effects, and result in a high Type 2 error rate.  Second, if the unobserved (i.e., latent) network structure is truly correlated with the observed covariates, the unobserved structure that can be correlated with the observed variable may be attributed to the observed variable. Under this condition the latent space parameters are used to model sources of variation that cannot be attributed to the observed covariate, and inferences regarding the observed covariates remain subject to omitted variable bias, which leads to an inflated Type 1 inferential error rate.
